---
title: "Most harmful event-types in the U.S."
author: "Spongebob_77"
date: "8 November 2018"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Important note
Due to proxy related issues I couldn't make the upload to RPubs. Also I couldn't
install LaTeX on the company computer. That's why I upload the html doc.

## Synopsis

Based on the *Storm Data* from NOAA (U.S. National Oceanic and Atmospheric 
Administration's) the main event-types shall be discovered with respect to 
population health and the economic consequences in the United States. The 
dataset contains information within a timeframe from 1950 till 2011. Due to 
reasons of data quality only the years 1992 till 2011 are used to answer the 
upper question. It turns out that tornados and lightnings are most harmful with 
respect to population health while floods, flashfloods and tornados cause the 
most economic damage.

## Data Preprocessing

### Libreries

First we load the libraries that are needed in this analysis.

```{r, libraries, echo=TRUE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
```

### Read in Data

The data is in a *.csv.bz2* file which is read in here. The data can be 
downloaded on [Coursera](https://www.coursera.org/learn/reproducible-research/peer/OMZ37/course-project-2).

Then we check the dataset with respect to dimension, classes the variables have, 
and how the first few rows of the dataset look like.

```{r, loadData, echo=TRUE, cache=TRUE}
# Load the data
dat_orig <- read.csv("repdata%2Fdata%2FStormData.csv.bz2")
dat <- dat_orig

# check variables
dim(dat)
str(dat)
head(dat)

```

As we know now how the dataset looks like we can start with the preprocessing.

### Select years

First we have a look how sparse the data is. We might not consider data from the 
beginning of the tracking since it is written in the assignment that there was 
more data available from the later years.

```{r yearsConsider, echo=TRUE}
# extract year and save into new column "year"
dat$year <- as.numeric(str_sub(dat$BGN_DATE, -12, -8))

# plot the number of datapoints each year contains
no_datapoints_years <- dat %>% count(year)
no_datapoints_years
plot(x = no_datapoints_years$year, 
     y = no_datapoints_years$n,
     type = "l",
     main = "number of datapoints per year",
     xlab = "year", ylab = "# datapoints")

# only keep the last twenty years of the data base
dat <- dat %>% filter(year >= 1992)
```

The data reaches from 1950 till 2011. In the first years the data was very 
sparse. So we decided to take into account only the last 20 years (1992-2011)
in which period we consider the data base as sufficient.

### Select main events

The next step is to focus on the main event types - meaning the ones that 
occure the most. We assume that these events cause the most trouble with regard 
to population health and economic consequences. We take the variable 
*INJURIES* to determine the population health and *CROPDMG* to determine 
the economic consequences.

```{r relevantEvents, echo=TRUE}
# no. of types of events
length(unique(dat$EVTYPE))

# check number of occurences of types of events
no_occ_events <- as.data.frame(sort(table(dat$EVTYPE), decreasing = TRUE))
head(no_occ_events)

# plot the top 40 events
barplot(sort(table(dat$EVTYPE), decreasing = TRUE)[1:40],
        las = 2,
        main = "sorted frequencies of events between 1992 and 2011 (only top 40)",
        xlab = "event", ylab = "frequency")

# save types of events with occurrences >= 10k (main-events)
no_occ_events <- no_occ_events %>% filter(Freq >=10000)
main_events <- as.character(no_occ_events[, 1])
main_events

# keep only data containing main-events
dat <- dat %>% filter(EVTYPE %in% main_events)
dim(dat)

# only keep relevant variables with respect to health and consequences:
# envtype, injuries, cropdmg and year
rel_vars <- colnames(dat)[c(8, 24, 27, 38)]
dat_final <- dat[, rel_vars]
dim(dat_final)
```

It turned out that many of the events are written in different ways (e.g. 
*TSTM WIND* and *THUNDERSTORM WIND*). Since we only can make assumptions about 
which events could be unified it was decided to keep them seperated and to 
consider them as unequal events.

According to the upper barplot we decided to make the cut at 10k events since 
we only take a look at the events that occur the most. We end up with 12 events 
in total.

Now we have the final dataset *dat_final* which will be the base for the analysis.

### Formatting

In order to find out which types of events have the biggest effect on 
population and crop in the United States, we calculate the mean effects per 
event-type and year.

In the end we will be able to compare the means of all event-types over the 
years 1992 till 2011.

```{r formatting, echo=TRUE}
# summarize by event-type and year for injuries and crop damages (means)
avgEffects <- dat_final %>% group_by(EVTYPE, year) %>% 
                summarise(Injuries = round(mean(INJURIES), 3),
                          CropDamage = round(mean(CROPDMG), 3))

head(avgEffects)

# wide to long format for results plot
avgEffects <- avgEffects %>% tidyr::gather(effectType, meanEffect, Injuries:CropDamage)

head(avgEffects)
```

## Results

We produce a line plot splitted by crop damage (to detect economic consequences) 
and by injuries (to detect most harmful event-types for humans).

```{r resultsPlot, echo=TRUE}
# panel plot
avgEffects %>% 
  ggplot(aes(x=year, y=meanEffect, colour=factor(EVTYPE))) + 
  geom_line() +
  facet_grid(effectType ~ ., scales="free") +
  ggtitle("mean effects of top 12 event-types over last 20 years") + 
  theme(legend.title = element_blank())
```

If we first take a look at the means from the crop damage we can see that floods 
have been the main cause in the past 20 years - especially in the early 90's.
In the years 2005 to 2011 we discover that crop damages were result mainly of 
floods, flash floods and tornados. Hail is another event that constantly have 
economic consequences, but way less than the other three events. So **floods, flash floods and tornados can be considered the as the greatest economic threads**.

In 1998 many people in the US were injured as a consequence of a flood. This was 
the most harmful event between 1992 and 2011. However, in the past few years 
tornados (especially in 2011) and lightnings have caused many injured people
quite constantly. So **tornados and lightnings are the biggest threads to the people**.






